---
title: "Morning Warm-up Questsions"
author: "C Mull"
date: "Feb 2022"
output: html_document
---

#Morning warm-ups

The following sections are just some warm-up questions for us to do together each morning to get our brains working and refresh what we covered the previous day.


```{r}
# Load cetacean.csv
getwd()

data <- read.csv("Data/Cetaceans.csv")

#Subset data to only include those entries from Scotland

names(data)
class(data$Age)
str(data)

scots <- subset(data, Location == "Scotland")

#How many entries are in the database?

length(scots)
dim(scots)
str(scots)

#How many unique species are in the Scotland database?



#What is the range of the ages in the Scotland database?



#Subset the original data to only include those entries from Scotland that are also from the Phocoenaphocoena and greater than 8 years of age. 



#Subset the database to only include those entries that are not from Scotland



#Plot the age distribution of cetacean from Scotland and Spain in boxplots by species and location. 



#Plot the age distribution of cetaceans by Dolphin.ID from Scotland and from Spain. Use a dot plot, make the Spain points orange and the Scotland points blue. 


```



## libraries
```{r libraries, include=FALSE}
#how do I load the `tidyverse` package

```

## Data loading

```{r brain size data}
#how do I load my `abalone.csv` file and create a new dataframe?

data <- read.csv("Data/abalone.csv")
```

## Including Plots

```{r plotting}
#how do I plot the entire dataset for an exploratory check?
plot(data)

#how do I plot Whole Weight over age?
plot(WholeWeight ~ Rings, data = data)

#how do I change this plot to distinguish between males and females?
males <- subset(data, Sex == "M")
females <- subset(data, Sex == "F")

plot(WholeWeight ~ Rings, data = females, xlim=range(data$Rings), ylim=range(data$WholeWeight), pch=19, col=alpha("red", 0.25), cex=1)

points(WholeWeight ~ Rings, data = males, xlim=range(data$Rings), ylim=range(data$WholeWeight), pch=19, col=alpha("blue", 0.25), cex=1)

#how can I create separate plots for males and females using facets in ggplot?

ggplot(data, aes(Rings, WholeWeight)) +
  geom_point(aes(color=Sex)) +
  facet_grid(~Sex)
```


## Data Wrangling

```{r}
#load the shark and ray brain size data file as an object call bs


#Plot the entire data set for exploration and hypothesis generation


#Create a new column that is a combination of telencephalon size and medulla size


#Create a new database that summarizes `bs` and has columns that count the number of species and mean body mass per habitat type. Hint: use `group_by` `summarize`.




```


### Pipe challenge!

When working within the tidyverse you can combine packages (dplyr and ggplot). Using this idea can you create a plot of the relative difference in body mass over depth, with a different color for each superorder.

```{r}


```

