---
title: "Morning Warm-up Questsions"
author: "C Mull"
date: "Feb 2022"
output: html_document
---

#Morning warm-ups

The following sections are just some warm-up questions for us to do together each morning to get our brains working and refresh what we covered the previous day.


```{r}
# Load cetacean.csv
getwd()

data <- read.csv("Data/Cetaceans.csv")

#Subset data to only include those entries from Scotland

names(data)
class(data$Age)
str(data)

scots <- subset(data, Location == "Scotland")

#How many entries are in the database?

length(scots)
dim(scots)
str(scots)

#How many unique species are in the Scotland database?



#What is the range of the ages in the Scotland database?



#Subset the original data to only include those entries from Scotland that are also from the Phocoenaphocoena and greater than 8 years of age. 



#Subset the database to only include those entries that are not from Scotland



#Plot the age distribution of cetacean from Scotland and Spain in boxplots by species and location. 



#Plot the age distribution of cetaceans by Dolphin.ID from Scotland and from Spain. Use a dot plot, make the Spain points orange and the Scotland points blue. 


```



## libraries
```{r libraries, include=FALSE}
#how do I load the `tidyverse` package

```

## Data loading

```{r brain size data}
#how do I load my `abalone.csv` file and create a new dataframe?

data <- read.csv("Data/abalone.csv")
```

## Including Plots

```{r plotting}
#how do I plot the entire dataset for an exploratory check?
plot(data)

#how do I plot Whole Weight over age?
plot(WholeWeight ~ Rings, data = data)

#how do I change this plot to distinguish between males and females?
males <- subset(data, Sex == "M")
females <- subset(data, Sex == "F")

plot(WholeWeight ~ Rings, data = females, xlim=range(data$Rings), ylim=range(data$WholeWeight), pch=19, col=alpha("red", 0.25), cex=1)

points(WholeWeight ~ Rings, data = males, xlim=range(data$Rings), ylim=range(data$WholeWeight), pch=19, col=alpha("blue", 0.25), cex=1)

#how can I create separate plots for males and females using facets in ggplot?

ggplot(data, aes(Rings, WholeWeight)) +
  geom_point(aes(color=Sex)) +
  facet_grid(~Sex)
```


## Data Wrangling

```{r}
#load the shark and ray brain size file as an object call bs

bs <- read.csv("Data/brain.data.csv")


#Plot the entire data set for exploration and hypothesis generation

plot(bs)

plot(bs)[,c("telencephalon":"body.mass")]

summary(bs)

#Create a new column that is a combination of telencephalon size and medulla size

transmute(bs, tel.med = telencephalon + medulla) 


#Create a new database that summarizes `bs` and has columns that count the number of species and mean body mass per habitat type. Hint: use `group_by` `summarize`.

new.dat <- bs %>%
  group_by(habitat) %>%
  summarise(numspec = length(species),
            mean.body = mean(body.mass))


```


### Pipe challenge!

When working within the tidyverse you can combine packages (dplyr and ggplot). Using this idea can you create a plot of the relative difference in body mass over depth, with a different color for each superorder.

group_by

summarise mean body

ungroup

mutate relative body

plot rel body/depth

across the various superorder

```{r}
bs %>%
  group_by(superorder) %>%
  mutate(body.mass = log10(body.mass)) %>%
  mutate(mean.body = mean(body.mass)) %>%
  mutate(rel.body = body.mass - mean.body) %>%
  group_by(.) %>%
  ggplot(aes(x = depth, y = rel.body)) +
  geom_point(aes(color=superorder)) +
  facet_wrap(~class)

```


## Welcome back

```{r}

# Load in the shark brain data set `brain.data.csv` data set

bd <- read.csv("Data/brain.data.csv")

# How do we examine the dataset to see the dimensions, number and names of variables, and the classes of each variable?

str(bd)
names(bd)
glimpse(bd)

# plot telencephalon over brain.mass

plot(bd$telencephalon ~ bd$brain.mass)

ggplot(bd, aes(brain.mass, telencephalon)) +
  geom_point()

# Recreate the plot but with habitat type signified by color

ggplot(bd, aes(brain.mass, telencephalon)) +
  geom_point(aes(color=habitat))

# Remove all of the deep water species and recreate the plot but with reproductive mode signified by color

bd %>%
  filter(habitat != "Deep") %>%
  ggplot(aes(brain.mass, telencephalon)) +
           geom_point(aes(color=reproductive.mode))

# Create a character vector that is every level of class in the `brain.data` set combined
#HINT use str_

unique(bd$class)

x <- str_c("Batoids" ,  "Sharks",    "Chimaeras", sep = " ")

# Just to be silly can you replace all of the vowels with question marks

str_replace_all(x,"[aeiou]","$")

# If we have untidy data where a single row contains multiple observations, which is common with large entered data sets, how would we reshape this into a tidy format? HINT see tidyr::table4a for a simple example.

tidyr::table4a

pivot_longer()
```


## Last Day...you're almost an expert!

```{r}

# load in the abalone dataset

ab <- read.csv("Data/abalone.csv")
str(ab)

# Which is more strongly correlated: Length and Height or Whole Weight and Shucked Weight?

cor.test(ab$Length, ab$Height)
cor.test(ab$WholeWeight, ab$ShuckedWeight)

# Run a linear regression of Shucked Weight over Length:

ab.mod <- lm(ShuckedWeight ~ Length, data = ab)
summary(ab.mod)
# What is the slope?
1.66

# What is the intercept?
-0.51

# What are the p-values telling us?

They are not 0.

# Does this regression vary by Sex? How would you check? 

ab.mod2 <- lm(ShuckedWeight ~ Length*Sex, data = ab)
summary(ab.mod2)


# Plot Shucked Weight over Length and color by sex:

plot(ShuckedWeight ~ Length, data = ab, pch=19, 
     col=factor(ab$Sex))
abline(ab.mod2$coefficients[1], ab.mod2$coefficients[2])
abline(ab.mod2$coefficients[1] + ab.mod2$coefficients[3],ab.mod2$coefficients[2] + ab.mod2$coefficients[5], col="red")
abline(ab.mod2$coefficients[1] + ab.mod2$coefficients[4],ab.mod2$coefficients[2] + ab.mod2$coefficients[6], col="green")

ggplot(ab, aes(Length, ShuckedWeight)) +
  geom_point(aes(color=Sex)) +
  geom_abline(intercept = ab.mod2$coefficients[1], slope =ab.mod2$coefficients[2])  +
  geom_abline(intercept = -0.74+0.09, slope = 2.05-0.13)

```

