---
title: "Rworkshop_multipanel plots"
author: "S Anderson and C Mull"
date: "24/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Multipanel Plots

This lesson was taken directly from the stats ninja [Sean Anderson](https://seananderson.ca/courses/11-multipanel/multipanel.pdf). See hsi website and follow his work for some amazing macroecology and analysis. He taught most of us R so all the credit for this lesson goes to him!

Often times our graphics in papers will contain mutliple panels to conserve space or to show information we want coupled together. All of our plotting so far has been single panel, but now we want to put several together! Sean lay's out 5 main approaches for doing this:

  1. By hand: full control but time consuming and not reproducible.
  2. `grid graphics` with packages like `lattice` or `ggplot2`. Like we have discussed before these are great for quick and easy exploratory plotting. But there is less user control so these might not be ideal for final figures for papers.
  3. `par(mfrow)`  he simplest method in base graphics. Works well for simple grid layouts where each panel is the same size. We actually ran some of these in our first plotting lesson, but they can only get so complex.
  4. `layout()` in addition to what you can do with par(mfrow), layout() lets you combine panels.
  5. `split.screen()` Lets you specify the co-ordinates of your panels. Panels no longer have to be simple ratios of each other.
  
The best method all depends on what works best for you as there is no right or wrong answer when it comes to making final plots...except maybe powerpoint.

# Layers of complexity

`par(mfrow)`, `layout()`, and `split.screen()` are all capable of basic equal-sized-panel grid layouts. If you think of creating a small multiple layout in R to be like putting screws into a wall: par(mfrow) would be the equivalent of grabbing your Leatherman to hang one picture frame — it’s all you need and it’s fast. layout() would be the equivalent of hunting around for a proper screwdriver to hang a bunch of picture frames. split.screen would be the equivalent of finding and plugging in your power drill — more of a hassle to set up, but much more powerful in the end. Don’t grab a tool that’s more complex than it needs to be, but don’t try and build
a house with a Leatherman.

# Questions to ask yourself when making a plot

  1. What comparison do I want to emphasize?
  2. How can I use order to enhance the comparison?
  3. Is this a series of plots or does the grid layout matter? (facet_wrap vs. facet_grid in ggplot2 terminology)
  4. What’s a reasonable number of panels to show? Everything? A sample?
  5. Which axes can I fix and which need to vary? Would a log transformation be appropriate and allow the axes to be combined?
  6. What chart junk can I remove?
  7. What’s important in my plots and what necessary but less-important elements do I want to de-emphasize?
  8. Can I make it all smaller and increase the information density without detract- ing from readability? (Almost always, yes.)
  9. If the layout is complicated, have I drawn it out on paper first?
  
# Margin space

Extra margins are usually wasted space and a break in the comparisons between panels. You will almost always want to shrink your margins. Set your margins for each panel with mar and your outer margins with `oma`. If all the axes can be shared then set `mar = c(0,0,0,0)`. These numbers refer to the space on the bottom, left, top, and right. Then you can use `par(oma)` to set your outer margins to create the necessary space for axes. If your content won’t show up in the outer margins, you’ll need to set par(xpd = NA).


# Ways to iterate through your data

Common approaches are to use a for loop with subsetting or an apply function. You could also manually make all your plots, but unless you were only making a few plots you wouldn’t do that would you? A favourite approach of mine is to use `d_ply()` from the `plyr` package. This takes a data frame, splits it up, does something with it (plots it), but doesn’t return a value. NOTE: plyr has been updated to `dplyr` but we could perform the same tasks by applying a function to our dataframe and creating data or objects we would like to plot.


# Basic multipanel layouts with par(mfrow)

For most basic grid layouts, `par(mfcol)` or `par(mfrow)` are your simplest option. `mfrow` plots row by row and `mfcol` plots column by column. mfrow is therefore likely the most commonly used option. You’re going to give `mfrow` vector of length two corresponding to the number of rows followed by the number of columns.

Let’s try a basic example with 2 rows and 3 columns:

```{r basic multipanel mfrow}
par(mfrow = c(2, 3))
par(cex = 0.6)
par(mar = c(3, 3, 0, 0), oma = c(1, 1, 1, 1))
for (i in 1:6) {
 plot(1, 1, type = "n")
 mtext(letters[i], side = 3, line = -1, adj = 0.1, cex = 0.6)
 }

```


We can eliminate the redundant axes, remove margin space, and reduce the emphasis on the structural (non-data) elements of the figure. These are some of the frequent “tricks” you can use to create a basic multipanel layout that will focus the reader’s attention on trends in the data. If you aren’t familiar with an option for `par()`, look up the help: `?par`.

```{r}
par(mfrow = c(2, 3))
 par(cex = 0.6)
 par(mar = c(0, 0, 0, 0), oma = c(4, 4, 0.5, 0.5))
 par(tcl = -0.25)
 par(mgp = c(2, 0.6, 0))
 for (i in 1:6) {
 plot(1, axes = FALSE, type = "n")
mtext(letters[i], side = 3, line = -1, adj = 0.1, cex = 0.6,
 col = "grey40")
 if (i %in% c(4, 5, 6))
 axis(1, col = "grey40", col.axis = "grey20", at = seq(0.6,
 1.2, 0.2))
 if (i %in% c(1, 4))
 axis(2, col = "grey40", col.axis = "grey20", at = seq(0.6,
 1.2, 0.2))
 box(col = "grey60")
 }
 mtext("x axis", side = 1, outer = TRUE, cex = 0.7, line = 2.2,
 col = "grey20")
 mtext("y axis", side = 2, outer = TRUE, cex = 0.7, line = 2.2,
 col = "grey20")
```


## Multipanel Plots with `layout`

Say you wanted to make a figure with one wide panel on top and two smaller panels underneath. We can’t do that with par(mfrow), now can we? This is where `layout()` becomes useful. It takes a matrix and turns it into a layout. The shape of the matrix corresponds to individual cells. The numbers in the matrix correspond to the order of the plots. Cells with the same number represent a single panel. I often find it easiest to create a matrix with rbind() and a series of vectors representing
the rows.

Let’s generate a layout like the previous example. First I will use the `layout.show()` command to see how the figure will look. The numbers correspond to the order that the panels will be plotted in.

```{r layout}

m <- rbind(c(1, 1), c(2, 3))
print(m)

layout(m)
layout.show(3)

layout(m)
par(mar = c(3, 3, 0, 0))
for (i in 1:3) plot(1, 1, type = "n")

```

This can be used to create some very complex multipanel plots (see pdf).

## Multipanel plots with `split.screen`

What if you want different panels to be different sizes and the sizes don’t correspond to some simple ratio that you could divide up with `layout()`? One common use for this is to keep the unit scales in different panels the same, without using extra margin space. `split.screen()` is only limited by your imagination and ability to figure out the co-ordinates of your plots.

Split screen takes a set of vectors or a matrix to specify the layout. I typically use the matrix notation. In this case each row describes a screen with values for left, right, bottom, and top.1 0 corresponds to the left or bottom of the plot. 1 corresponds
to the top or right of the plot. Sometimes it’s easiest to figure out this matrix in a spreadsheet before and then read it in as a .csv file. 

First let’s generate a “simple” two-by-two layout with `split.screen()`.


```{r split screen}

m <- rbind(c(0.1, 0.55, 0.55, 1), c(0.55, 1, 0.55, 1), c(0.1, 0.55, 0.1, 0.55), c(0.55, 1, 0.1, 0.55))
split.screen(m)
for(i in 1:4) {
screen(i)
par(mar = c(0, 0, 0, 0))
plot(1, axes = FALSE, type = "n")
box()
if(i %in% c(3, 4)) axis(1)
if(i %in% c(1, 3)) axis(2)
}
close.screen(all.screens = TRUE)

```


Here’s an example of setting up the panels to arbitrary sizes so that the units are equal on both x-axes. I figured out these values in a spreadsheet first.

```{r}

split.screen(rbind(c(0.1,0.292,0.1, 0.98), c(0.312, 0.95, 0.1, 0.98)))
screen(1)
par(mar = c(0, 0, 0, 0))
plot(1:30, rnorm(30), xaxs = "i", ylim = c(-3, 3), xaxt = "n")
axis(1, at = seq(0, 30, 20))

screen(2)
par(mar = c(0, 0, 0, 0))
plot(1:100, rnorm(100), xaxs = "i", ylim = c(-3, 3), yaxt = "n", col = "red")
close.screen(all.screens = TRUE)
```


Multipanel plotting takes a bit of planning but can be really powerful and by doing the coding in R you can easily reproduce these plots even when your data, code, or analysis gets updated.
